Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-20 appendix.tex
optimization section of appendix and marginal plot stuff

>>>>

\subsection*{Empirical validation}

We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}'$ as well as any optimization method available.

We take various safeguards to ensure convergence and stability of the procedure.
Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x$ or $y$ are either $0$ or $1$.
As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x, y < 1$ and we try to evaluate the case of $x'=y'=w'=1$ we have a likelihood of exactly zero.
Worse still, in Table~\ref{tab:likelihoods} for the Farris topology, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $y'=1$ will similarly yield a zero likelihood.
In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
We sidestep these computational issues by restricting ourselves to the region $x,y\in [10^{-3},1-10^{-3}]^2$.

We initialize our optimization procedure at the true branch parameter vector $\{x, y, y\}$.
We posit that $\hat{w}'$ is equal to one for regions of $x$ and $y$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial value is small, so we perform two maximizations, one with $w'$ fixed and one where it is estimated.
We take the value of $\hat{w}'$ with the larger likelihood as our estimate.

### 2017-11-20 joint_inf.tex
optimization section of appendix and marginal plot stuff

The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation using numerical optimization.
Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.

The estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001-marginal.svg --out-pkl-name figures/w-hat-empirical-001-marginal.pkl --n-jobs 16 --marginal
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001-marginal.svg --in-pkl-name figures/w-hat-empirical-001-marginal.pkl
% if fit already
% !! TODO: update with finer-grained plot; this is a placeholder as the global optimization runs...
\includegraphics[width=\textwidth]{w-hat-empirical-001-marginal}
\caption[short]{Marginal inference optimizing \eqref{eq:marginal_likelihood}}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --out-pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-001}
\caption[short]{Joint inference optimizing \eqref{eq:profile_likelihood}}
\end{subfigure}
\caption{
    %das: note to update caption
    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using differential evolution \cite{Storn1997}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

>>>>

The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, described in the Appendix.
Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.

The case of marginal inference in as expected, where $\hat{w}'$ is equal to $y$ regardless of the value of $x$.
For joint inference, the estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001-marginal.svg --out-pkl-name figures/w-hat-empirical-001-marginal.pkl --n-jobs 16 --marginal
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001-marginal.svg --in-pkl-name figures/w-hat-empirical-001-marginal.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-001-marginal}
\caption[short]{Marginal inference optimizing \eqref{eq:marginal_likelihood}}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --out-pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-001}
\caption[short]{Joint inference optimizing \eqref{eq:profile_likelihood}}
\end{subfigure}
\caption{
    %das: note to update caption
    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using basin-hopping \cite{Wales1997}.
}
\label{fig:bl-general-inconsistency}
\end{figure}
